\section{Related Work}
\label{sec:related-work}

% Pig
Pig \cite{olston_pig_2008-1} is a framework for writing jobs for the Hadoop platform using an imperative domain specific language called Pig latin. Pig latin's restricted interface allows the Pig system to apply relational optimizations that include operator rewrites, early projection and early filtering to achieve good performance. 
It has extensive support for relational operations and allows the user to choose between different join implementations with varying performance characteristics. Pig latin users need to learn a new language which is not the case with frameworks such as Hadoop, Hive and Crunch. It does not include user defined functions, the user needs to define them externally in another language, which will often prevent optimizations as Pig can not analyze those. Pig latin is not Turing complete as it does not include control structures itself.
The language is not statically type checked so runtime failures are common and time consuming. Also, pure Java approaches benefit from a rich ecosystem of productivity tools. 

Even though \tool also adopts a domain specific approach, it is deeply embedded in Scala, Turing complete and allows the user to easily define functions which do not disable the optimizations. Currently \tool does not have support for many relational optimizations. However, it includes compiler optimizations and it is well extensible.
%In \tool we also adopt the domain specific approach but we give the user much more freedom. User defined functions can be written directly in \tool and compiler and relational optimizations are not disabled by them. \tool is syntactically not a new language and it is embedded into the Scala language which gives it access to most of the productivity tools. Although we currently do not support many relational optimizations in \tool we apply aggressive compiler optimizations jointly with relational ones. \tool can be easily extended with additional optimizations without modifying the framework it self.

% Steno 
Steno \cite{murray_steno:_2011} is a .NET library that, through runtime code generation, effectively removes abstraction overhead of the LINQ programming model. It removes all iterator calls inside LINQ queries and provides significant performance gains in CPU intensive jobs on Dryad/LINQ. Queries that use Steno do not limit the generality of the programming model but optimizations like code motion and early projection are not possible. 
\tool also removes excess iterator calls from the code but during operation fusion it enables other optimizations, especially when combined with early projection. The drawback of \tool is that it is slightly limited in generality.

% Manimal and HadoopToSQL
Manimal \cite{jahani_automatic_2011} and HadoopToSQL \cite{iu_hadooptosql:_2010} perform static byte code analysis on Hadoop jobs to infer different program properties that can be mapped to relational optimizations. They both use the inferred program properties to build indexes and achieve much more efficient data access patterns. Manimal can additionally organize the data into columnar storage. These approaches are limited by the incomplete program knowledge which is lost by compilation and runtime determined functions. They both do not restrict the programming model at all. 
\tool shares the idea of providing code generality to these approaches. However, it currently does not include data indexing schemes which could enable big performance improvements. We believe that the full type and program information available in \tool will enable us to build better data indexing schemes for a larger set of user programs.
